---
title: "task4_markdown"
author: "Kirsty Anderson"
date: "2023-04-11"
output: html_document
---

1. Introduction
--------------------------------------------------------------------------------
2. Assumptions
- the dimensions of all 3 tables is different

3. Cleaning the Data
4. Analysis Answers

load packages and read data
```{r}
library(tidyverse)
library(here)
```

```{r}
candy_full <- read_csv(here::here("clean_data/candy_cleaned.csv"))
```


Q1: What is the total number of candy ratings given across the three years. (Number of candy ratings, not the         number of raters. Don’t count missing values)

A: There were 772352 ratings

```{r}
candy_full %>% 
  select(candy, rating) %>% 
  filter(!is.na(rating)) %>% 
  nrow()
```
Q2: What was the average age of people who are going out trick or treating?
A:
check what answers are in column going out
```{r}
candy_full %>% 
  distinct(going_out)
```
Filter for "yes" to going out
# check what answers are in age col
```{r}
candy_full %>% 
  distinct(age)
```

only include age that is numner

```{r}
candy_full_num_age <- candy_full %>% 
  filter(!is.numeric(age))

candy_full %>% 
  distinct(age)
```

5. Conclusions and notes

