---
title: "task4_markdown"
author: "Kirsty Anderson"
date: "2023-04-11"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

# 1. Introduction

This dataset is a combination of 3 sets of answers to 3 consecutive annual surveys from 2015, 2016, and 2017 about Halloween candy.
The survey differs slightly each year, but in general it asks the users' age, gender, location, if they plan on "trick - or - treating", and then goes on to ask the user their opinion on various types of Halloween candy.
It also asks the users opinion on various other topics.

I have cleaned the dataset in the file task4_cleaning_script.R, and then performed analyses on the data in this file.

------------------------------------------------------------------------

# 2. Assumptions

-   I have made some assumptions in regards to column names across the 3 sets e.g. bonkers(2015) = bonkers_the_candy and not bonkers_the_boardgame (2016 & 2017). See table for transformations: \|

    +:----------------------------------------------------------------------:+:------------------------------------------------------------------------:+:-------------------------------------------------------------------------------:+:-------------------------:+
    | ### **2015**                                                           | ### 2016                                                                 | ### 2017                                                                        | ### Change                |
    +------------------------------------------------------------------------+--------------------------------------------------------------------------+---------------------------------------------------------------------------------+---------------------------+
    | box_o\_raisins                                                         | boxo_raisins                                                             | boxo_raisins                                                                    | 2015 to match 16/17       |
    +------------------------------------------------------------------------+--------------------------------------------------------------------------+---------------------------------------------------------------------------------+---------------------------+
    | bonkers                                                                | bonkers_the_candy/ bonkers_the_board_game                                | bonkers_the_candy/\                                                             | 2015 to bonkers_the_candy |
    |                                                                        |                                                                          | bonkers_the_board_game                                                          |                           |
    +------------------------------------------------------------------------+--------------------------------------------------------------------------+---------------------------------------------------------------------------------+---------------------------+
    | sweetums                                                               | sweetums_a\_friend_to_diabetes                                           | sweetums_a\_friend_to_diabetes                                                  | 2015 to match 16/17       |
    +------------------------------------------------------------------------+--------------------------------------------------------------------------+---------------------------------------------------------------------------------+---------------------------+
    | anonymous_brown_globs_that_come_in_black_and_orange_wrappers           | anonymous_brown_globs_that_come_in_black_and_orange_wrappers             | anonymous_brown_globs_that_come_in_black_and_orange_wrappers_a\_k_a\_mary_janes | 2017 to match 15/16       |
    +------------------------------------------------------------------------+--------------------------------------------------------------------------+---------------------------------------------------------------------------------+---------------------------+
    | that_dress_that_went_viral_early_this_year_when_i\_first_saw_it_it_was | that_dress_that_went_viral_a\_few_years_back_when_i\_first_saw_it_it_was | q10_dress                                                                       | all to "viral_dress"      |
    +------------------------------------------------------------------------+--------------------------------------------------------------------------+---------------------------------------------------------------------------------+---------------------------+
    | which_day_do_you_prefer_friday_or_sunday                               | which_day_do_you_prefer_friday_or_sunday                                 | q11_day                                                                         | 2017 to match 15/16       |
    +------------------------------------------------------------------------+--------------------------------------------------------------------------+---------------------------------------------------------------------------------+---------------------------+
    | x100_grand_bar                                                         | x100_grand_bar                                                           | 100_grand_bar                                                                   | 2017 to match 15/16       |
    +------------------------------------------------------------------------+--------------------------------------------------------------------------+---------------------------------------------------------------------------------+---------------------------+

    : Candy Column Name Transformations

-   I have assumed that any variations of usa e.g. U.S or U.S of A is the same as usa

-   I have taken the age column at face-value - i.e. have not removed any impossible age values such as 200587.

------------------------------------------------------------------------

# 3. Cleaning the Data

Load packages and Read Data

```{r}
library(tidyverse)
library(here)
library(stringr)
```

-   I first transformed the columns to snake_case using the janitor::clean_names function.
-   I then dealt with the column headers. Across the 3 sets, the column names differed and it was obvious that some columns were the same across differnt years, just with slightly different naming e.g. the column titled "how_old_are_you" in the 2015 and 2016 sets would be the same as "age" in the 2017 set. I used the recode() function to change these, the full transformation can be seen in the script but for example:

```{r}
# candy_2015_clean <- candy_2015_clean %>% 
#  rename("going_out" = "are_you_going_actually_going_trick_or_treating_yourself", 
#         "age"       = "how_old_are_you")
```

-   I also spent some time matching the candy columns from different data sets e.g.

```{r}
# candy_2015_clean <- candy_2015_clean %>% 
#   rename("boxo_raisins"                  = "box_o_raisins",
#          "bonkers_the_candy"             = "bonkers",
#          "sweetums_a_friend_to_diabetes" = "sweetums",
#          "viral_dress" = "that_dress_that_went_viral_early_this_year_when_i_first_saw_it_it_was")

```

-   I knew that I was going to join the 3 datasets together, and wanted to preserve the year each observation was recorded, I added a constant value column containing the year to the beginning of each dataset e.g.:

```{r}
# candy_2015_clean <- candy_2015_clean %>% 
#   add_column(year = 2015,.after = "going_out")
```

-   I then boung the 3 datasets, for any missing values I filled with NA using the plyr::rbind.fill() function:

```{r}
# candy_clean_bound <- plyr::rbind.fill(candy_2015_clean, 
#                                 candy_2016_clean, 
#                                 candy_2017_clean)
```

-   I then converted the dataset to long format to make it tidy, creating a column for candy and a column for rating. I left all non-candy columns in wide format.

```{r}
# candy_clean_long <- pivot_longer(candy_clean_bound,
#                                  cols = c("butterfinger": "york_peppermint_patties", 
#                                           "necco_wafers", 
#                                           "sea_salt_flavored_stuff_probably_chocolate_since_this_is_the_it_flavor_of_the_year",
#                                           "bonkers_the_board_game":"whatchamacallit_bars",
#                                           "green_party_m_ms":"take_5"),
#                                            names_to = "candy",
#                                            values_to = "rating")  %>% 
#                     relocate(candy, .after = going_out) %>% 
#                     relocate(rating, .after = candy)
```

-   Now that the columns were reasonably clean and tidy, I dealt with some of the "dirty" values, specifically the country and age column.
-   The country column was free text input, so it contained lots of entry styles from the same country.
-   In the future, I would like to create a more efficient formula to recode this, but for now I recoded it manually as well as I could, for example:

```{r}
# candy_clean_long <- candy_clean_long %>% 
#   mutate(country = recode(country,
#                           "'merica" = "usa",
#                           "america" = "usa",
#                           )) 

```

-   The age column contained string values, so couldn't be converted to a numerical format.
-   I replaced any strings with NA, and any infinite values as NA

```{r}
# candy_clean_long$age[!grepl("^-?[0-9]+(\\.[0-9]+)?$", candy_clean_long$age)] <- NA 
# candy_clean_long$age <- as.numeric(candy_clean_long$age) # (2) convert the column to numeric
# candy_clean_long$age <- round(candy_clean_long$age, digits = 0) # (3) round age to nearest integer 
# candy_clean_long$age[sapply(candy_clean_long$age, is.infinite)] <- NA # replace infinite with NA
```

-   I then wrote this cleaned, transformed data to csv and saved it into the cleaned_data folder.

------------------------------------------------------------------------

# 4. Analysis Answers

Load Data

```{r}
candy_full <- read_csv(here::here("clean_data/candy_cleaned.csv"))
```

Q1: What is the total number of candy ratings given across the three years.

```         
(Number of candy ratings, not the number of raters. Don’t count missing values)
```

```{r}
# count number of rows that have valid candy ratings

candy_full %>% 
  select(candy, rating) %>% 
  filter(!is.na(rating)) %>% 
  nrow()

```

A: There were 772352 ratings

Q2: What was the average age of people who are going out trick or treating?

```{r}
# 1: check what answers are in column "going_out" & check age col is numerical

candy_full %>% 
  distinct(going_out)


class(candy_full$age)
 
```

```{r}
# 2: Filter for "yes" to going out and find mean age, discount NA values
candy_full %>% 
  filter(going_out == "Yes") %>% 
  summarise(avg_age = round(mean(age, na.rm = TRUE), digits = 0))
```

A: The average age for people that were going out is 253.

Q3: What was the average age of people who are not going trick or treating?

```{r}
# As above, filter for "no" to going out and find mean age, discount NA values
candy_full %>% 
  filter(going_out == "No") %>% 
  summarise(avg_age = round(mean(age, na.rm = TRUE), digits = 0))
```

A: The average age for people that were going_out is 40.

Q4: For each of joy, despair and meh, which candy bar received the most of these ratings?

```{r}
# Create a vector containing each rating type
rating_options <- c("JOY", "DESPAIR", "MEH")

# Create a loop that filters the DF for each rating, then counts number of candies
for (vote in rating_options){
  max_candy <-  candy_full %>% 
                filter(rating == vote) %>% 
                count(candy) %>% 
                arrange(desc(n)) %>% 
                head(1)
phrase <- paste("The candy '",max_candy[1,1],"' received the most",vote,"ratings, with",max_candy[1,2],"ratings.")
print(phrase)
}

```

Since "any_full_sized_candy_bar" doesn't provide much information, repeat code above and discount this to get second top choice:

```{r}
rating_options <- c("JOY", "DESPAIR", "MEH")


for (vote in rating_options){
  max_candy <-  candy_full %>% 
                filter(rating == vote) %>%
                filter(candy != "any_full_sized_candy_bar") %>% 
                count(candy) %>% 
                arrange(desc(n)) %>% 
                head(1)
phrase <- paste("The candy '",max_candy[1,1],"' received the most",vote,"ratings, with",max_candy[1,2],"ratings.")
print(phrase)
}
```

Q5: How many people rated Starburst as despair?

```{r}
# Filter candy column for starburst and rating column for despair, then count rows.
candy_full %>% 
  filter(candy == "starburst") %>% 
  filter(rating == "DESPAIR") %>% 
  summarise(count = n())
```

A: 1990 people rated startburst with despair

For the next three questions, count despair as -1, joy as +1, and meh as 0.

```{r}
#recode each rating to its new number, save this to a new column called "num_rating"
candy_full <- candy_full %>% 
      mutate(num_rating = recode(rating,
                              "JOY" = 1,
                              "DESPAIR" = -1,
                              "MEH" = 0)) %>% 
      relocate(num_rating, .after = rating)

```

Q6: What was the most popular candy bar by this rating system for each gender in the dataset ?

```{r}
# create a vector containing all possible gender categories, called "gender_option"
gender_option <- c("Female", "Male", "Other", NA)

# For each gender option, if it is not NA, then filter the DF for the selected gender, summarise the total points for each candy and print the highest one
for (gender_cat in gender_option){
  if (!is.na(gender_cat)){
    max_candy <- candy_full %>% 
    filter(gender == gender_cat) %>% 
    group_by(candy) %>% 
    summarise(points = sum(num_rating, na.rm = TRUE)) %>% 
    arrange(desc(points)) %>% 
    head(1)
  
phrase <- paste("When gender was",gender_cat, "the top rated candy bar was",max_candy[1,1],"which scored",max_candy[1,2],"points")
    
print(phrase)  
  } 
  # if the gender is NA, do the same thing but filter for gender is NA
  else {
    max_candy <- candy_full %>% 
    filter(is.na(gender)) %>% 
    group_by(candy) %>% 
    summarise(points = sum(num_rating, na.rm = TRUE)) %>% 
    arrange(desc(points)) %>% 
    head(1)
  
phrase <- paste("When gender was NA the top rated candy bar was",max_candy[1,1],"which scored",max_candy[1,2],"points")
    
print(phrase)  
    
  }}
 
```

As in Question 4: "any_full_sized_candy_bar" isn't really a candy choice, repeat code above and discount this to get second top choice:

```{r}

gender_option <- c("Female", "Male", "Other", NA)


for (gender_cat in gender_option){
  if (!is.na(gender_cat)){
    max_candy <- candy_full %>% 
      filter(gender == gender_cat) %>% 
      filter(candy != "any_full_sized_candy_bar") %>% 
      group_by(candy) %>% 
      summarise(points = sum(num_rating, na.rm = TRUE)) %>% 
      arrange(desc(points)) %>% 
      head(1)
    
    phrase <- paste("When gender was",gender_cat, "the top rated candy bar was",max_candy[1,1],"which scored",max_candy[1,2],"points")
    
    print(phrase)  
  } 
  
  else {
    max_candy <- candy_full %>% 
      filter(is.na(gender)) %>% 
      filter(candy != "any_full_sized_candy_bar") %>% 
      group_by(candy) %>% 
      summarise(points = sum(num_rating, na.rm = TRUE)) %>% 
      arrange(desc(points)) %>% 
      head(1)
    
    phrase <- paste("When gender was NA the top rated candy bar was",max_candy[1,1],"which scored",max_candy[1,2],"points")
    
    print(phrase)  
    
  }}

```

Rerun again but also remove cash_or_other_forms_of_legal_tender!

```{r}

gender_option <- c("Female", "Male", "Other", NA)


for (gender_cat in gender_option){
  if (!is.na(gender_cat)){
    max_candy <- candy_full %>% 
      filter(gender == gender_cat) %>% 
      filter(candy != "any_full_sized_candy_bar") %>% 
      filter(candy != "cash_or_other_forms_of_legal_tender") %>% 
      group_by(candy) %>% 
      summarise(points = sum(num_rating, na.rm = TRUE)) %>% 
      arrange(desc(points)) %>% 
      head(1)
    
    phrase <- paste("When gender was",gender_cat, "the top rated candy bar was",max_candy[1,1],"which scored",max_candy[1,2],"points")
    
    print(phrase)  
  } 
  
  else {
    max_candy <- candy_full %>% 
      filter(is.na(gender)) %>% 
      filter(candy != "any_full_sized_candy_bar") %>% 
      filter(candy != "cash_or_other_forms_of_legal_tender") %>% 
      group_by(candy) %>% 
      summarise(points = sum(num_rating, na.rm = TRUE)) %>% 
      arrange(desc(points)) %>% 
      head(1)
    
    phrase <- paste("When gender was NA the top rated candy bar was",max_candy[1,1],"which scored",max_candy[1,2],"points")
    
    print(phrase)  
    
  }}
```

Q7.
What was the most popular candy bar in each year?

```{r}
# create a vector containing all possible years, called "year_option"
year_option <- c(2015, 2016, 2017)

# For each year option, filter the DF for the selected year, summarise the total points for each candy and print the highest one
for(year_cat in year_option){
  max_year <- candy_full %>% 
              filter(year == year_cat) %>% 
              group_by(candy) %>% 
              summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
              arrange(desc(total)) %>% 
              head(1)
  phrase <- paste("In",year_cat,"the most popular candy bar was '",max_year[1,1],"', which scored",max_year[1,2],"points.")
  print(phrase)
}

```

Run again but omit "any_full_sized_candy_bar" and "cash_or_other_forms_of_legal_tender"

```{r}


year_option <- c(2015, 2016, 2017)


for(year_cat in year_option){
  max_year <- candy_full %>% 
              filter(year == year_cat) %>% 
              filter(candy != "any_full_sized_candy_bar") %>%
              filter(candy != "cash_or_other_forms_of_legal_tender") %>% 
              group_by(candy) %>% 
              summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
              arrange(desc(total)) %>% 
              head(1)
  phrase <- paste("In",year_cat,"the most popular candy bar was '",max_year[1,1],"', which scored",max_year[1,2],"points.")
  print(phrase)
}

```

Q8.
What was the most popular candy bar by this rating for people in US, Canada, UK, and all other countries?

```{r}
# 1: Get list of countries to see names
candy_full %>% 
  distinct(country)
```

1: USA:

```{r}
# Filter where country contains the string "usa", group by candy, count ratings for each candy and find the top
# Check filter first:

#  candy_full %>% 
#  filter(str_detect(country, "usa")) %>% 
#  filter(!str_detect(country, "not")) %>% 
#  distinct(country)


candy_full %>% 
  filter(str_detect(country, "usa")) %>% 
  filter(!str_detect(country, "not")) %>% 
  group_by(candy) %>% 
  summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
  arrange(desc(total))


```

A: Any_full_sized_candy_bar scored the highest in the USA at 2158 points however does not technically answer the question.
The highest actual candy bar was Reese's Peanut Butter Cups that scored 1971 points.

2: Canada

```{r}
# Check filter:

# candy_full %>% 
#   filter(str_detect(country, "canada" )) %>% 
#   filter(!str_detect(country, "not")) %>%
#   distinct(country)
  
  
  candy_full %>% 
  filter(str_detect(country, "canada" )) %>% 
  filter(!str_detect(country, "not")) %>%
  group_by(candy) %>% 
  summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
  arrange(desc(total))
```

A: Any_full_sized_candy_bar scored the highest in Canada at 251 points however does not technically answer the question.
The highest actual candy bar was Kit-Kat that scored 229 points.

3: UK:

```{r}
# Check filter:

#  candy_full %>% 
#     filter(str_detect(country, "uk|u.k.|united|kingdom|scotland|ireland|england|wales" )) %>% 
#     filter(!str_detect(country, "not|states|soviet")) %>%
#     distinct(country)
  
  
  candy_full %>% 
  filter(str_detect(country, "uk|u.k.|united|kingdom|scotland|ireland|england|wales" )) %>% 
  filter(!str_detect(country, "not|states|soviet")) %>%
  group_by(candy) %>% 
  summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
  arrange(desc(total))

```

A: The highest scoring in the UK was cash, but the highest scoring actual candy was Lindt truffles at 35 points.

4: Other Countries

```{r}
# Check Filter:

# candy_full %>% 
#     filter(!str_detect(country, "canada|uk|u.k.|united|kingdom|scotland|ireland|england|wales|usa" )) %>%
#     distinct(country)

candy_full %>% 
 filter(!str_detect(country, "canada|uk|u.k.|united|kingdom|scotland|ireland|england|wales|usa" )) %>% 
  group_by(candy) %>% 
  summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
  arrange(desc(total))
```

A: The highest scoring elsewhere was again full sized candy bars at 101 points, but the highest scoring actual candy was Kit-Kat at 77 points.

------------------------------------------------------------------------

5.  Conclusions and notes

There is a lot more cleaning of this dataset that I would like to return to carry out, e.g. creating a better function to create a uniform country column, and finding out if the numbers in the country column relate to anything.
I also feel there is more analysis that I could carry out such as finding trends amongst age and gender, however as the survey was very open it is hard to get accurate data findings - e.g. the age column ranges from 1 to 200587.
