---
title: "task4_markdown"
author: "Kirsty Anderson"
date: "2023-04-11"
output: html_document
---

1. Introduction
--------------------------------------------------------------------------------
2. Assumptions
- the dimensions of all 3 tables is different

3. Cleaning the Data
4. Analysis Answers

load packages and read data
```{r}
library(tidyverse)
library(here)
library(stringr)
```

```{r}
candy_full <- read_csv(here::here("clean_data/candy_cleaned.csv"))

```


Q1: What is the total number of candy ratings given across the three years. (Number of candy ratings, not the         number of raters. Don’t count missing values)

A: There were 772352 ratings

```{r}
candy_full %>% 
  select(candy, rating) %>% 
  filter(!is.na(rating)) %>% 
  nrow()
```
Q2: What was the average age of people who are going out trick or treating?
A:
check what answers are in column going out & check class of age col
```{r}
candy_full %>% 
  distinct(going_out)
 
class(candy_full$age)

```
Filter for "yes" to going out
# check what answers are in age col
```{r}
candy_full %>% 
  distinct(age)
```



```{r}
candy_full %>% 
  filter(going_out == "Yes") %>% 
  summarise(avg_age = round(mean(age, na.rm = TRUE), digits = 0))
```
A: The average age for people that were going_out is 253.

Q3: What was the average age of people who are not going trick or treating?

```{r}
candy_full %>% 
  filter(going_out == "No") %>% 
  summarise(avg_age = round(mean(age, na.rm = TRUE), digits = 0))
```
A: The average age for people that were going_out is 40.


Q4: For each of joy, despair and meh, which candy bar received the most of these ratings?

```{r}
rating_options <- c("JOY", "DESPAIR", "MEH")

for (vote in rating_options){
  max_candy <-  candy_full %>% 
                filter(rating == vote) %>% 
                count(candy) %>% 
                arrange(desc(n)) %>% 
                head(1)
phrase <- paste("The candy '",max_candy[1,1],"' received the most",vote,"ratings, with",max_candy[1,2],"ratings.")
print(phrase)
}

```


Q5: How many people rated Starburst as despair?

```{r}
candy_full %>% 
  filter(candy == "starburst") %>% 
  filter(rating == "DESPAIR") %>% 
  summarise(count = n())
```
A: 1990 people rated startburst with despair

For the next three questions, count despair as -1, joy as +1, and meh as 0.

```{r}
#recode each rating to its new number, save this to a new column called "num_rating"
candy_full <- candy_full %>% 
  mutate(num_rating = recode(rating,
                             "JOY" = 1,
                             "DESPAIR" = -1,
                             "MEH" = 0)) %>% 
  relocate(num_rating, .after = rating)

```

Q6 What was the most popular candy bar by this rating system for each gender in the dataset ?


```{r}
# create a vector containing all possible gender categories, called "gender_option"
gender_option <- c("Female", "Male", NA)

# For each gender option, if it is not NA, then filter the df for the selected gender, summarise the total points for each candy and print the highest one
for (gender_cat in gender_option){
  if (!is.na(gender_cat)){
    max_candy <- candy_full %>% 
    filter(gender == gender_cat) %>% 
    group_by(candy) %>% 
    summarise(points = sum(num_rating, na.rm = TRUE)) %>% 
    arrange(desc(points)) %>% 
    head(1)
  
phrase <- paste("When gender was",gender_cat, "the top rated candy bar was",max_candy[1,1],"which scored",max_candy[1,2],"points")
    
print(phrase)  
  } 
  # if the gender is na, do the same thing but filter for gender is NA
  else {
    max_candy <- candy_full %>% 
    filter(is.na(gender)) %>% 
    group_by(candy) %>% 
    summarise(points = sum(num_rating, na.rm = TRUE)) %>% 
    arrange(desc(points)) %>% 
    head(1)
  
phrase <- paste("When gender was NA the top rated candy bar was",max_candy[1,1],"which scored",max_candy[1,2],"points")
    
print(phrase)  
    
  }}
 
```

Q7. What was the most popular candy bar in each year?

```{r}
year_option <- c(2015, 2016, 2017)

for(year_cat in year_option){
  max_year <- candy_full %>% 
              filter(year == year_cat) %>% 
              group_by(candy) %>% 
              summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
              arrange(desc(total)) %>% 
              head(1)
  phrase <- paste("In",year_cat,"the most popular candy bar was '",max_year[1,1],"', which scored",max_year[1,2],"points.")
  print(phrase)
}

```

Q8. 
What was the most popular candy bar by this rating for people in US, Canada, UK, and all other countries?
get list of countries
```{r}
candy_full %>% 
  distinct(country)
```

```{r}
#USA:
candy_full %>% 
              filter(country == "[usa]") %>% 
              group_by(candy) %>% 
              summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
              arrange(desc(total)) %>% 
              head(1)

```

```{r}
# USA 
candy_full %>% 
  filter(str_detect(country, "usa")) %>% 
  filter(!str_detect(country, "not")) %>% 
  group_by(candy) %>% 
  summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
  arrange(desc(total))


```
A: any_full_sized_candy_bar which scored 2158 points.

```{r}
# UK: check filter:

candy_full %>% 
  filter(str_detect(country, "uk|u.k.|united|kingdom|scotland|ireland|england|wales" )) %>% 
  filter(!str_detect(country, "not|states|soviet")) %>%
  distinct(country)
  
  
  candy_full %>% 
  filter(str_detect(country, "uk|u.k.|united|kingdom|scotland|ireland|england|wales" )) %>% 
  filter(!str_detect(country, "not|states|soviet")) %>%
  group_by(candy) %>% 
  summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
  arrange(desc(total))

```
```{r}
# Canada: check filter:

candy_full %>% 
  filter(str_detect(country, "canada" )) %>% 
  filter(!str_detect(country, "not")) %>%
  distinct(country)
  
  
  candy_full %>% 
  filter(str_detect(country, "canada" )) %>% 
  filter(!str_detect(country, "not")) %>%
  group_by(candy) %>% 
  summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
  arrange(desc(total))
```
```{r}
# all other countries

candy_full %>% 
  filter(!str_detect(country, "canada|uk|u.k.|united|kingdom|scotland|ireland|england|wales|usa" )) %>%
  distinct(country)

candy_full %>% 
 filter(!str_detect(country, "canada|uk|u.k.|united|kingdom|scotland|ireland|england|wales|usa" )) %>% 
  group_by(candy) %>% 
  summarise(total = sum(num_rating, na.rm = TRUE)) %>% 
  arrange(desc(total))
```


5. Conclusions and notes


