---
title: "task1_markdown"
author: "Kirsty Anderson"
date: "2023-04-10"
output: html_document
---
1. Introduction
--------------------------------------------------------------------------------
2. Assumptions
- Sports are measured in metres or seconds depending on the sport.
3. Cleaning the Data
- change row names to column called "athlete" using dplyr's row_name_to_column
- change to long format
- rearrange
- convert competition to string?
- count NAs
4. Analysis Answers
5. Conclusions and notes


--------------------------------------------------------------------------------
# Load libraries

```{r}
library(tidyverse)
library(here)
```

# Load in cleaned data

```{r}
decathlon <- read_csv(here("clean_data/decathlon_cleaned.csv"))

```
# Analysis Questions 

Q1: Who had the longest long jump seen in the data?

- First, check what sports are listed to see if there are >1 long_jump categories:
```{r}
decathlon %>% distinct(sport)
```
- There is only one long_jump variable

```{r}
# select the long_jump entry with the highest "score".
decathlon %>% 
  filter(sport == "long_jump") %>% 
  arrange(desc(score_metres_seconds)) %>% 
  head(1)
```
A: Clay had the longest long jump at 7.96 metres.
--------------------------------------------------------------------------------

Q2: What was the average 100m time in each competition?

```{r}
# Get the average 100m time in each competition
decathlon %>% 
  filter(sport == "x100m") %>% 
  group_by(competition) %>% 
  summarise(average_100m = round(mean(score_metres_seconds), digits =2)) 

  
```
A: The average 100m time in the Decastar competition was 11.18 seconds, in the OlympicG competition it was       10.92 seconds.
--------------------------------------------------------------------------------

Q3: Who had the highest total points across both competitions?

```{r}
# For each athlete, count the total number of points, then select the highest
decathlon %>% 
  group_by(athlete) %>% 
  summarise(total_points = sum(points)) %>% 
  arrange(desc(total_points)) %>% 
  head(1)
 
```
A: The athlete with the highest total points across both competitions was Sebrle.

---------------------------------------------------------------------------------

Q4: What was the shot-put scores for the top three competitors in each competition?

A: 
```{r}
decathlon %>% 
  group_by(competition) %>% 
  filter(sport == "shot_put" & rank %in% c(1, 2, 3)) %>% 
  arrange(desc(points))
  
```

--------------------------------------------------------------------------------

Q5: What was the average points for competitors who ran the 400m in less than 50 seconds vs. those than ran 400m in more than 50 seconds?

```{r}
decathlon %>% 
  filter(sport == "x400m" & score_metres_seconds < 50.00) %>% 
  summarise(avg_less_50s = round(mean(points), digits = 0)) %>% 
  pull(avg_less_50s)

decathlon %>% 
  filter(sport == "x400m" & score_metres_seconds > 50.00) %>% 
  summarise(avg_more_50s = round(mean(points), digits = 0)) %>% 
  pull(avg_more_50s)

```
A: For competititors who ran the 400m in <50 seconds, the average points was 8120. For competitors who did it in >50 seconds, the average points was 7727.


